---
title: "Gender Pay Gaps"
format: html
editor: visual
Author: "Dency Claris Thomas"
---

## The data-set used

Install the stevedata package in R and use the gss_wages data set. We can read the documentation about the data.

## Format

A data frame with 11 variables:

`year`

:   the survey year

`realrinc`

:   the respondent's base income (in constant 1986 USD)

`age`

:   the respondent's age in years

`occ10`

:   respondent's occupation code (2010)

`occrecode`

:   recode of the occupation code into one of 11 main categories

`prestg10`

:   respondent's occupational prestige score (2010)

`childs`

:   number of children (0-8)

`wrkstat`

:   the work status of the respondent (full-time, part-time, temporarily not working, unemployed (laid off), retired, school, housekeeper, other)

`gender`

:   respondent's gender (male or female)

`educcat`

:   respondent's degree level (Less Than High School, High School, Junior College, Bachelor, or Graduate)

`maritalcat`

:   respondent's marital status (Married, Widowed, Divorced, Separated, Never Married)

```{r}
library(stevedata)

library(tidyverse)
```

```{r}
gss_wages|> glimpse()
```

```{r}
wage_data <- as_tibble(gss_wages)

wage_data

wage_data_filtered <- wage_data[complete.cases(wage_data), ]
wage_data_filtered
```

### Bar plot to visualize the average income by gender

```{r}
# Average income by gender

avg_income_gender <- wage_data_filtered |>
  group_by(gender) |>
  summarise(avg_income = mean(realrinc, na.rm = TRUE))
avg_income_gender

# Bar plot
ggplot(avg_income_gender, aes(x = gender, y = avg_income, fill = gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Income by Gender", x = "Gender", y = "Average Income (USD)") +
  theme_minimal()
```

### Bar plot to visualize the average income by gender (Filtered with interquartile range)

```{r}

# Function to filter data within the interquartile range
filter_iqr <- function(x) {
  lower <- quantile(x, 0.25, na.rm = TRUE)
  upper <- quantile(x, 0.75, na.rm = TRUE)
  return(between(x, lower, upper))
}

# Filter data within the interquartile range
filtered_data <- wage_data_filtered |>
  group_by(gender) |>
  filter(filter_iqr(realrinc))

# Calculate the average income after filtering
avg_income_filtered <- filtered_data |>
  group_by(gender) |>
  summarise(avg_income = mean(realrinc, na.rm = TRUE))
avg_income_filtered

# Bar plot
ggplot(avg_income_filtered, aes(x = gender, y = avg_income, fill = gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Income by Gender (Filtered within Interquartile Range)",
       x = "Gender",
       y = "Average Income (USD)",
       fill = "Gender") +
  theme_minimal()

```

### Time Trends in Gender Pay Gap:

The resulting chart will show how the average income has changed over the survey years for both genders, allowing us to observe trends in the gender pay gap over the survey years.

```{r}
# Convert 'year' to a factor for better visualization
wage_data$year <- as.factor(wage_data$year)

# Calculate average income for each gender and year
avg_income_gender_year <- wage_data |>
  group_by(year, gender) |>
  summarise(avg_income = mean(realrinc, na.rm = TRUE))
  avg_income_gender_year
  
# Create a point chart
ggplot(avg_income_gender_year, aes(x = avg_income, y = year, color = gender)) +
  geom_point() +
  labs(title = "Time Trends in Gender Pay Gap",
       x = "Average Income",
       y = "Year",
       color = "Gender") +
  theme_minimal() 
```

### Box Plots of Income by Education Level:

\- Using box plots to compare the distribution of income for different education levels (educcat) separately for males and females.

```{r}
# Convert 'gender' to a factor for better visualization
wage_data$gender <- as.factor(wage_data$gender)

# Create box plots
ggplot(wage_data_filtered, aes(x = realrinc , y = educcat, fill = gender)) +
  geom_boxplot() +
  labs(title = "Box Plots of Income by Education Level",
       x = "Income (realrinc)",
       y = "Education Level",
       fill = "Gender") +
  theme_minimal() 
```

### Scatter Plot of Age vs. Income:

\- Visualizing the correlation between age and income using a scatter plot, with different colors for males and females.

```{r}
# Create a scatter plot with alpha transparency
ggplot(wage_data, aes(x = age, y = realrinc, color = gender)) +
  geom_point(alpha = 0.5, na.rm = TRUE) +
  labs(title = "Scatter Plot of Age vs. Income",
       x = "Age",
       y = "Income (realrinc)",
       color = "Gender") +
  theme_minimal()
```

### Grouped Bar Plot of Income vs Work Status by Gender

```{r}

# Grouped Bar Plot
ggplot(wage_data_filtered, aes(x = wrkstat, y = realrinc, fill = gender)) +
  geom_bar(stat = "summary", fun = "mean", position = "dodge") +
  labs(title = "Income vs Work Status by Gender",
       x = "Work Status",
       y = "Mean Income (realrinc)",
       fill = "Gender") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

### Marital Status vs Income by Gender

```{r}

# Box Plot
ggplot(wage_data_filtered, aes(x = maritalcat, y = realrinc, fill = gender)) +
  geom_boxplot() +
  labs(title = "Income vs Marital Status by Gender",
       x = "Marital Status",
       y = "Income (realrinc)",
       fill = "Gender") +
  theme_minimal() 

```

### Distribution of Marital Status and Work Status by Gender

```{r}
# Stacked Bar Plot
ggplot(wage_data_filtered, aes(x = maritalcat, fill = wrkstat)) +
  geom_bar(position = "stack") +
  facet_wrap(~gender, scales = 'free_y') +
  labs(title = "Distribution of Marital Status and Work Status by Gender",
       x = "Marital Status",
       fill = "Work Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Average Income by Number of Children

```{r}
# Remove rows with NA values
wage_data_filtered <- wage_data[complete.cases(wage_data[, c('childs', 'realrinc', 'gender', 'maritalcat')]), ]

# Create Grouped Bar Plot
ggplot(wage_data_filtered, aes(x = factor(childs), y = realrinc, fill = gender)) +
  geom_bar(stat = "summary", fun = "mean", position = "dodge") +
  labs(title = "Average Income by Number of Children",
       x = "Number of Children",
       y = "Average Income (realrinc)",
       fill = "Gender") +
  theme_minimal()

```

### Distribution of Income Levels by Number of Children

```{r}
# Define income categories (you can adjust these based on your data)
income_categories <- cut(wage_data_filtered$realrinc, breaks = c(0, 20000, 40000, 60000, 80000, Inf),
                          labels = c("0-20k", "20k-40k", "40k-60k", "60k-80k", "80k+"), include.lowest = TRUE)

# Create Stacked Bar Plot
ggplot(wage_data_filtered, aes(x = factor(childs), fill = income_categories)) +
  geom_bar(position = "fill") +
  facet_wrap(~gender, scales = 'free_y') +
  labs(title = "Distribution of Income Levels by Number of Children",
       x = "Number of Children",
       fill = "Income Level") +
  theme_minimal()

```

### Correlation between age, income and child

```{r}
library(corrplot)

# Select numerical variables for correlation analysis
numerical_vars <- wage_data_filtered |>
  select(age, realrinc, childs)

# Calculate correlation matrix
cor_matrix <- cor(numerical_vars, use = "complete.obs")

# Create correlation heatmap
corrplot(cor_matrix, method = "color", addCoef.col = "black", tl.col = "black")

```
